### 2024-06-08 journal/analysis.tex
some edits from final proofreading
some final edits from proofreading

\section{Analysis}
We now analyze the number of steps and the number of \op{CAS} instructions performed by operations.

@@

\begin{lemma}\label{dSearchTime}
The search that \op{FindResponse}$(b,i)$ does at line \ref{FRb} to find the index $b_e$ of the block in the root containing the $e$th \enqueue\ takes $O(\log (\var{root}.\fld{blocks}[b_e].\size + \var{root}.\fld{blocks}[b-1].\size))$ steps.
\end{lemma}
\begin{proof}
Let the blocks in the root be $B_1, \ldots, B_\ell$.
The doubling search for $b_e$ takes $O(\log (b-b_e))$ steps,
so we prove $b - b_e \leq 2 \cdot B_{b_e}.\size + B_{b-1}.\size + 1$.
If $b \leq b_e+1$, then this is trivial, so assume  $b>b_e+1$.
%
As shown in Lemma \ref{linearCorrect}, the \dequeue\ that calls \op{FindResponse} is in $B_b$ and is supposed to return an \enqueue\ in $B_{b_e}$.
Thus, there can be at most $B_{b_e}.\size$ \dequeues\ in 
$D(B_{b_e+1}) \cdots D(B_{b-1})$; otherwise in the sequential execution $L$,
all elements enqueued before the end of
$E(B_{b_e})$ would be dequeued before $D(B_b)$. 
Furthermore, by Lemma \ref{sizeCorrectness}, the size of the queue  after the prefix of $L$ corresponding to 
$B_1,\ldots,B_{b-1}$  is 
$B_{b-1}.\size \geq B_{b_e}.size + |E(B_{b_e+1})\cdots E(B_{b-1})| - |D(B_{b_e+1})\cdots D(B_{b-1})|$.
Thus, $|E(B_{b_e+1})\cdots E(B_{b-1})| \leq B_{b-1}.\size + |D(B_{b_e+1})\cdots D(B_{b-1})| \leq B_{b-1}.\size + B_{b_e}.\size$.
So, the total number of operations in $B_{b_e+1}, \ldots, B_{b-1}$ is at most
$B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
Each of these $b-1-b_e$ blocks contains at least one operation, by Corollary \ref{blockNotEmpty}.
So, $b-1-b_e \leq B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
\end{proof}

The following lemma helps bound the \rev{number of steps \op{GetEnqueue} takes}.
\begin{lemma}\label{blockSize}
Each block $B$ in each node contains at most one operation of each process.
If $c$ is the execution's maximum point contention, $B$ has at most $c$ direct~subblocks.
\end{lemma}
\begin{proof}
Suppose $B$ contains an operation of process $p$.
Let $op$ be the earliest operation by $p$ contained in $B$.
When $op$ terminates, $op$ is contained in $B$ by Lemma \ref{lem::appendExactlyOnce}.
$B$ cannot contain any later operations by $p$, since $B$ is created before
those operations are invoked.

>>>>

\section{Analysis}
\label{sec-analysis}

We now analyze the number of steps and the number of \op{CAS} instructions performed by operations.

@@

\begin{lemma}\label{dSearchTime}
The search that \op{FindResponse}$(b,i)$ does at line \ref{FRb} to find the index $b_e$ of the block in the root containing the $e$th \enqueue\ takes $O(\log (\var{root}.\fld{blocks}[b_e].\size + \var{root}.\fld{blocks}[b-1].\size))$ steps.
\end{lemma}
\begin{proof}
Let the blocks in the root be $B_1, \ldots, B_\ell$.
The doubling search for $b_e$ takes $O(\log (b-b_e))$ steps,
so we prove $b - b_e \leq 2 \cdot B_{b_e}.\size + B_{b-1}.\size + 1$.
If $b \leq b_e+1$, then this is trivial, so assume  $b>b_e+1$.
%
As shown in Lemma \ref{linearCorrect}, the \dequeue\ that calls \op{FindResponse} is in $B_b$ and is supposed to return an \enqueue\ in $B_{b_e}$.
Thus, there can be at most $B_{b_e}.\size$ \dequeues\ in 
$D(B_{b_e+1}) \cdots D(B_{b-1})$; otherwise in the sequential execution defined by $L$,
all elements enqueued before the end of
$E(B_{b_e})$ would be dequeued before $D(B_b)$. 
Furthermore, by Lemma \ref{sizeCorrectness}, the size of the queue  after the prefix of $L$ corresponding to 
$B_1,\ldots,B_{b-1}$  is 
$B_{b-1}.\size \geq B_{b_e}.size + |E(B_{b_e+1})\cdots E(B_{b-1})| - |D(B_{b_e+1})\cdots D(B_{b-1})|$.
Thus, $|E(B_{b_e+1})\cdots E(B_{b-1})| \leq B_{b-1}.\size + |D(B_{b_e+1})\cdots D(B_{b-1})| \leq B_{b-1}.\size + B_{b_e}.\size$.
So, the total number of operations in $B_{b_e+1}, \ldots, B_{b-1}$ is at most
$B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
Each of these $b-1-b_e$ blocks contains at least one operation, by Corollary \ref{blockNotEmpty}.
So, $b-1-b_e \leq B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
\end{proof}

The following lemma helps bound the \rev{number of steps \op{GetEnqueue} takes}.
\begin{lemma}\label{blockSize}
Each block $B$ in each node contains at most one operation of each process.
If $c$ is the execution's maximum point contention, $B$ has at most $c$ direct~subblocks.
\end{lemma}
\begin{proof}
Suppose $B$ contains an operation of process $p$.
Let $op$ be the earliest operation by $p$ contained in $B$.
When $op$ terminates, $op$ is contained in $B$, by Lemma \ref{lem::appendExactlyOnce}.
$B$ cannot contain any later operations by $p$, since $B$ is created before
those operations are invoked.

### 2024-06-08 journal/conclusion.tex
some final edits from proofreading

It would be interesting to close the gap that remains between our queue, which takes $O(\log^2 p + \log q)$ steps per operation,
and Jayanti, Tarjan and Boix-Adser\`{a}'s $\Omega(\log p)$ lower bound \cite{JTB19}.
For the more relaxed bag data structure the gap is larger between the  $\Omega(\min(c,\log\log p))$ lower bound \cite{DBLP:conf/opodis/AttiyaF17} and our upper bound of $O(\log^2 p + \log q)$.
Could the complexity for either queues or bags be made polylogarithmic in $p$ while being independent of the size $q$ of the data structure?

@@

Building on the work described in this paper, Asbell and Ruppert \cite{AR23} have
designed a doubly-ended queue with polylogarithmic amortized step
complexity, which also yields a stack as a special case.  
This required a substantially different representation of the data stored in the ordering tree.
Whether the ordering tree could also be used to obtain a priority queue with polylogarithmic step complexity
remains an open question.

>>>>

It would be interesting to close the gap that remains between our queue, which takes $O(\log^2 p + \log q)$ steps per operation,
and Jayanti, Tarjan and Boix-Adser\`{a}'s $\Omega(\log p)$ lower bound~\cite{JTB19}.
For the more relaxed bag data structure the gap is larger between the  $\Omega(\min(c,\log\log p))$ lower bound \cite{DBLP:conf/opodis/AttiyaF17} and our upper bound of $O(\log^2 p + \log q)$.
Could the complexity for either queues or bags be made polylogarithmic in $p$ while being independent of the size $q$ of the data structure?

@@

Building on the work described in this paper, Asbell and Ruppert \cite{AR23} have
designed a doubly-ended queue with polylogarithmic amortized step
complexity, which also yields a stack as a special case.  
This required a substantially different representation of the data stored in the ordering tree.
Whether the ordering tree could also be used to obtain a priority queue with polylogarithmic step complexity
remains an open question.
\rev{The technique of cooperatively propagating information up a tree with a double refresh
at each node was also used recently to obtain lock-free augmented search trees \cite{FR24}.}

### 2024-06-08 journal/correctness-GC.tex
some final edits from proofreading

If line \ref{casGC} of a \op{Refresh} modifies $\var{v}.\fld{blocks}$, then \var{v} is an internal node.
After reading $T$ from $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC},
the \op{Refresh} then creates the  block \var{B},
and calls \op{AddBlock}($v,T,B$) to create $T'$.
Line \ref{createIndexGC} sets $\var{B}.\var{index} = \max(I)+1$.
The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $B$ to it.
So the claim is satisfied.
\end{proof}

@@

Next, we prove an analogue of Lemma \ref{successfulRefresh}.
We say a node \var{v} \emph{contains} an operation
if some block containing the operation has previously appeared in the RBT \var{v}.\fld{block}
(even if the block has been removed from the RBT by a subsequent \op{Split} during garbage collection).

@@

Suppose $R$'s call to \op{CreateBlock} returns a new block $B$
that is installed in $\var{v}.\fld{blocks}$
by $R$'s \op{CAS} at line \ref{casGC}.
The \op{CreateBlock} set $B.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
By Corollary \ref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
By the definition of subblocks, some block in \var{v} contains $B_\ell$ as a subblock
and therefore \var{v} contains~$op$.

Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOPGC}.
Let $h$ be the maximum \fld{index} in $T$ plus 1.
By reasoning identical to the last paragraph of Lemma \ref{successfulRefresh}'s proof,
it follows from the fact that $\var{num\sub{enq}}+\var{num\sub{deq}}=0$ at line \ref{testEmptyGC}
that the blocks $\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..B.\eright]$
contain no operations.
By Corollary \ref{blockNotEmpty}, each block contains at least one operation, so
these ranges must be empty, and $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{B}.\eleft\geq i$.
This implies that the block $B_\ell$ containing $op$ is a subblock of some block that has appeared
in \var{v}'s RBT, so $op$ is contained in \var{v}.
\end{proof}

This allows us to show that a double \op{Refresh} propagates operations up the tree, 
as in Lemma \ref{lem::doubleRefresh}.

@@

We remark that Invariant \ref{inv::finished} guarantees that GC keeps one block that is finished and discards all blocks with smaller indices.
This is because the first unfinished block may still be traversed by an operation in the future,
and when examining that block the operation may need information from the preceding block.  For example,
when \op{FindResponse} is called on a block with index $b$, line \ref{FRnumenqGC} looks up the block with
index $b-1$.

\begin{lemma}\label{blocknotfound}
If a \op{Dequeue} operation fails to find a block in an RBT,
%during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
then it has been propagated to the root and its result has been written in the \fld{response} field of the leaf block it created.
If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
\end{lemma}
\begin{proof}
Any block that GC removes from an RBT is finished, by Invariant \ref{inv::finished}, so if an 
\op{Enqueue} or \op{Dequeue} fails to find a block while it is propagating itself up to
the root (for example, during the \op{CreateBlock} routine), then a block containing
the operation itself has been removed from a RBT, so the operation has propagated to the root,
by the definition of finished.
Moreover, by Invariant \ref{inv::finished}, if the operation is a \dequeue, then its result is in its \fld{response} field.

@@

Next, we bound the number of operations in \var{v}'s blocks that are retained when $A$  sets
$T'\assign\op{Split}(T,s)$.
Since $T$ was read before $t$, any operation in $T$ is either in progress at $t$ or has been propagated to the root before $t$, by Lemma \ref{lem::appendExactlyOnce}.
Thus, there are at most $p$ operations in $T$ that do not appear in $L_1\cdot L_2$.
All the rest of the operations in blocks of $T'$ have been propagated to blocks $m..\ell$ of the root.
By Lemma \ref{blockSize}, There are at most $p$ operations in block $m$ of the root 
and we showed above that there are at most $3q_{max} + 3p$ in blocks 
$m+1..\ell$ of the root.
Thus, there are at most $3q_{max}+5p$ operations in blocks of $T'$.
Since each block is non-empty by Corollary \ref{blockNotEmpty}, $T'$ contains at most $3q_{max}+5p$ blocks, and one more 
block is inserted before $A$ sets \var{v}'s \fld{blocks} to the resulting RBT.
\end{proof}

\begin{corollary}\label{RBTbound}
At all times, the size of a node's RBT is $O(q_{max}+p+G)$. 
\end{corollary}
\begin{proof}
Each update to a node's RBT adds at most one block to it, increasing its maximum \fld{index} by 1.
Thus, there are at most $G$ updates since the last time its maximum \fld{index} was a multiple of $G$.
The claim follows from Lemma \ref{boundAfterGC}.
\end{proof}

@@

Now we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}
and show that the amortized number of GC steps each operation performs is also $O(\log p\log(p+q_{max}))$.
If a process does GC in a call to \op{AddBlock}($\var{v},T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th GC phase on \var{v}}.

>>>>

If line \ref{casGC} of a \op{Refresh} modifies $\var{v}.\fld{blocks}$, then \var{v} is an internal node.
After reading $T$ from $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC},
the \op{Refresh} then creates the block \var{B},
and calls \op{AddBlock}($v,T,B$) to create $T'$.
Line \ref{createIndexGC} sets $\var{B}.\var{index} = \max(I)+1$.
The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT \rev{at line \ref{splitGC}}
and then add $B$ to it \rev{at line \ref{GCend} or \ref{noGC-insert}}.
So the claim is satisfied.
\end{proof}

@@

Next, we prove an analogue of Lemma \ref{successfulRefresh}.
We say a node \var{v} \emph{contains} an operation
if some block containing the operation has previously appeared in the RBT \var{v}.\fld{blocks}
(even if the block has been removed from the RBT by a subsequent \op{Split} during garbage collection).

@@

Suppose $R$'s call to \op{CreateBlock} returns a new block $B$
that is installed in $\var{v}.\fld{blocks}$
by $R$'s \op{CAS} at line \ref{casGC}.
The \op{CreateBlock} set $B.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
By Corollary \ref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than~$i$.
By the definition of subblocks, some block in \var{v} contains $B_\ell$ as a subblock
and therefore \var{v} contains~$op$.

Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOPGC}.
Let $h$ be the maximum \fld{index} in $T$ plus 1.
By reasoning identical to the last paragraph of Lemma \ref{successfulRefresh}'s proof,
it follows from the fact that $\var{num\sub{enq}}+\var{num\sub{deq}}=0$ at line \ref{testEmptyGC}
that the blocks $\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..B.\eright]$
contain no operations.
By Corollary~\ref{blockNotEmpty}, each block contains at least one operation, so
these ranges must be empty, and $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{B}.\eleft\geq i$.
This implies that the block $B_\ell$ containing $op$ is a subblock of some block that has appeared
in \var{v}'s RBT, so $op$ is contained in \var{v}.
\end{proof}

This allows us to show that a double \op{Refresh} propagates operations up the tree, 
as in Lemma~\ref{lem::doubleRefresh}.

@@

We remark that Invariant \ref{inv::finished} guarantees that GC keeps one block that is finished and discards all blocks with smaller indices.
This is because the first unfinished block may still be traversed by an operation in the future,
and when examining that block the operation may need information from the \emph{preceding block}.  For example,
when \op{FindResponse} is called on a block with index $b$, line \ref{FRnumenqGC} looks up the block with
index $b-1$.

\begin{lemma}\label{blocknotfound}
If a \op{Dequeue} operation fails to find a block in an RBT,
%during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
then \rev{the \dequeue} has been propagated to the root and its result has been written in the \fld{response} field of the leaf block it created.
If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
\end{lemma}
\begin{proof}
Any block that GC removes from an RBT is finished, by Invariant \ref{inv::finished}, so if an 
\op{Enqueue} or \op{Dequeue} fails to find a block while it is propagating itself up to
the root (for example, during the \op{CreateBlock} routine), then a block containing
the operation itself has been removed from a RBT, so the operation has propagated to the root,
by the definition of finished.
Moreover, by Invariant \ref{inv::finished}, if the operation is a \dequeue, then its result is in its \fld{response} field.

@@

Next, we bound the number of operations in \var{v}'s blocks that are retained when $A$  sets
$T'\assign\op{Split}(T,s)$.
Since $T$ was read before $t$, any operation in $T$ is either in progress at $t$ or has been propagated to the root before $t$, by Lemma~\ref{lem::appendExactlyOnce}.
Thus, there are at most $p$ operations in $T$ that do not appear in $L_1\cdot L_2$.
All the rest of the operations in blocks of $T'$ have been propagated to blocks $m..\ell$ of the root.
By Lemma \ref{blockSize}, There are at most $p$ operations in block $m$ of the root 
and we showed above that there are at most $3q_{max} + 3p$ in blocks 
$m+1..\ell$ of the root.
Thus, there are at most $3q_{max}+5p$ operations in blocks of $T'$.
Since each block is non-empty by Corollary \ref{blockNotEmpty}, $T'$ contains at most $3q_{max}+5p$ blocks, and one more 
block is inserted before $A$ sets \var{v}'s \fld{blocks} to the resulting RBT.
\end{proof}

\begin{corollary}\label{RBTbound}
At all times, the size of a node's RBT is $O(q_{max}+p+G)$. 
\end{corollary}
\begin{proof}
Each update to a node's RBT adds at most one block to it, increasing its maximum \fld{index} by~1.
Thus, there are at most $G$ updates since the last time its maximum \fld{index} was a multiple of $G$.
The claim follows from Lemma \ref{boundAfterGC}.
\end{proof}

@@

Now, we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}
and show that the amortized number of GC steps each operation performs is also $O(\log p\log(p+q_{max}))$.
If a process does GC in a call to \op{AddBlock}($\var{v},T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th GC phase on \var{v}}.
